### 2024-01-18 globals/metadata.tex
acm style
re-organising transducers and such

\title{A corrected version of Karhumäki’s theorem
       on polynomial $\Nat$-rational series}
\author{
    Lopez Aliaume
}
\date{\today}
\bibliography{globals/papers}

\newcommand{\acknowledge}{
    I would like to thank Gaëtan Douéneau-Tabot
    for mentionning this problem, checking 
    my counter-example, and supporting me in publishing
    this paper.
}

\newcommand{\makeabstract}{
    \begin{abstract}
        We propose a short counter-example to a
        previous result on commutative $\Nat$-rational series
        by Karhumäki, and provide a proof for
        the corrected statement.
        This opens the way to the decidability of
        $\Nat$ polyregular functions 
        among $\Rel$ porylegular functions, which is
        an ongoing research.
    \end{abstract}
}

>>>>

\title{
    Star-free $\Nat$-polyregular functions:
    the commutative case.
}
\author{
    Aliaume Lopez
}
\affiliation{%
  \institution{Univeristy of Warsaw}
  \city{Warsaw}
  \country{Poland}}
\email{ad.lopez@uw.edu.pl}

\date{\today}
\bibliography{globals/papers}

\newcommand{\acknowledge}{
    I would like to thank 
    Anonymous Author(s)
    %Gaëtan Douéneau-Tabot
    for mentioning this problem, checking 
    my counter-example, and supporting me in publishing
    this paper.
}

\newcommand{\makeabstract}{
    \begin{abstract}
        This paper answers two questions about 
        commutative unary output polyregular functions, called
        $\Nat$-polyregular functions, that forms a proper subset
        of $\Nat$-rational series.
        We show that given a $\Rel$-rational series,
        it is decidable whether it is computable via a commutative
        $\Nat$-polyregular function, and provide a counter-example
        to the theorem of Karhumäki that studied the same question
        in the case of polynomials.

        We also prove that it is decidable whether a commutative $\Nat$-polyregular
        function is star-free, by proving the stronger statement
        that star-free $\Rel$-polyregular functions that are
        $\Nat$-polyregular are computable using a star-free $\Nat$-polyregular function.
        This answers positively to previous conjectures on $\Nat$-polyregular functions,
        and the study of the commutative case seems to be useful
        towards a proof in the non-commutative case.
    \end{abstract}
}

\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

### 2024-01-18 parts/correction.tex
re-organising transducers and such

%! TeX program = xelatex
%! lang = en-US
\section{The Corrected Theorem}
\label{sec:proof}

The counter example provided by \cref{def:bad-polynomial} relies on the fact
that $\CoveredPoly$ is not stable under fixing indeterminates, while
\kl{$\Nat$-polyregular functions} are stable under the residual operation
$u^{-1} \colon f \mapsto f(u \cdot)$. In this section, we prove that closing
$\CorrectPoly$ under assignments of variable is enough to recover from
\cref{karh:thm}.

@@

\begin{lemma}
    \label{n-poly-combinatorics:lem}
    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
    There exists $\omega \in \Nat$
    such that for all $p \in \Nat$,
    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
    for all $u_1, \dots, u_p \in \Sigma^*$,
    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
    whose \kl{maximal monomials} are \kl{non-negative},
    and such that for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)
        \right)
        = P(X_1, \dots, X_p) \quad .
    \end{equation*}
\end{lemma}
\begin{proof}
    Let $\omega$ be an idempotent power for the finite monoid $M$,
    and
    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
    $f = \pi^\dagger$. That is, 
    $f(w)$ is the sum over all factorizations of $w$
    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.

@@

\begin{lemma}
    \label{n-rat-correct:lem}
    Let $P \in \Rel[X_1, \dots, X_p]$ be \kl{represented}
    by a \kl{$\Nat$-rational series}. Then,
    $P \in \CorrectPoly$.
\end{lemma}
\begin{proof}
    Let $f$ be the \kl{commutative}
    \kl{$\Nat$-rational series}
    with domain $\Sigma \defined \set{a_1, \dots, a_p}$
    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
    it is in fact a \kl{$\Nat$-polyregular function}
    (\cref{polynomial-rational-polyreg:fact}).
    Using \cref{n-poly-combinatorics:lem},
    there exists a number $\omega \in \Nat$
    and polynomial $Q$ with \kl{positive coefficients} for its
    \kl{maximal monomials} such that
    for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^p (a_i)^{\omega \times X_i}
        \right)
        = Q(X_1, \dots, X_p)
        = P(\omega X_1, \dots, \omega X_p) 
        \quad .
    \end{equation*}
    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
    as polynomials,
    and in particular that
    the \kl{maximal monomials} of 
    $P$ have \kl{positive coefficients}.

@@

Let us also assert that polynomials that have all coefficients in $\Nat$ are
\kl{$\Nat$-rational polynomials}. In particular, \kl{non-negative}
\kl{monomials} are \kl{$\Nat$-rational polynomials}, which will be a base case
in an upcoming induction proving \cref{lem:correct-to-n-rat}.

\begin{lemma}
    \label{n-poly-n-poly:example}
    Let $P \in \Nat[\vec{X}]$. Then, $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{lemma}
\begin{proof}
    We prove it by remarking that $\NPoly$
    is stable under non-negative linear combinations
    and contains the functions ...
\end{proof}

\AP The core of the upcoming \cref{lem:correct-to-n-rat} leverages a notion of
\kl{discrete derivative} to perform an induction on the \kl{maximal monomials},
and suitable notions of \kl{translations} of the domain of polynomials. Namely,
we write $\translate{K}$ for the \intro{translation function} that maps a
polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 - K, \dots,
X_n - K)$.

@@

In the actual statement of \cref{corrected-version:thm}, the representation of
polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
functions} to \kl{star-free $\Nat$-polyregular functions}. This has to be
compared with the characterization of \kl{star-free $\Rel$-polyregular
functions} in terms of polynomials \cite[Theorem??]{LOPEZ23b},
that we will re-introduce in \cref{star-free:sec}.

@@

The following two lemmas show how the \kl{discrete differentiation} operator
interacts with \kl{maximal monomials} with respect to the \kl{Hoare ordering}:
it extracts \emph{submaximal} monomials from a given polynomial.

@@

\begin{fact}
    Let $M,T$ be two \kl{monomials}, such that
    $T$ \kl{strictly divides} $M$.
    Then,
    $T$ \kl{divides} some \kl{maximal monomial}
    of $\Diff{K}{M}$.
\end{fact}

\begin{lemma}
    \label{derivation-covers:lemma}
    For all $P,Q \in \Rel[\vec{X}]$ that are non-constant,
    for all $K \in \Nat$,
    if $\MaximalMonomials(P) \hoarele \MaximalMonomials(Q)$,
    then
    $\MaximalMonomials(P) \hoareleq \MaximalMonomials(\Diff{K}{Q})$.
\end{lemma}

@@

While \cref{lem:correct-to-n-rat} provides effective conversion, it does not
explicitly state that the membership is decidable. However, it is not too
difficult to remark that $\CorrectPoly$, the proof scheme can be followed to
obtain decidability. For the sake of completeness and readability,
we extract this fact in the following lemma.

\begin{lemma}
    \label{decidability-correct:lem}
    Membership in $\CorrectPoly$ is decidable for
    polynomials in $\Rel[\vec{X}]$.
\end{lemma}
\begin{proof}
    Let $P \in \Rel[\vec{X}]$. We prove by induction on
    $\MaximalMonomials(P)$ that whether $P \in \CorrectPoly$
    is decidable.

>>>>

%! TeX program = xelatex
%! lang = en-US
\subsection{The Corrected Theorem}
\label{sec:proof}

The counter example provided by \cref{def:bad-polynomial} relies on the fact
that $\CoveredPoly$ is not stable under fixing indeterminates, while
\kl{$\Nat$-polyregular functions} are. In this section, we prove that closing
$\CorrectPoly$ under assignments of variable is enough to recover from
\cref{karh:thm}. Because the previous attempt to characterize this class of
polynomials did not hold, this part of the paper will try to be extremely
precise.

@@

\begin{lemma}
    \label{n-poly-combinatorics:lem}
    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
    There exists $\omega \in \Nat$
    such that for all $p \in \Nat$,
    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
    for all $u_1, \dots, u_p \in \Sigma^*$,
    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
    whose \kl{maximal monomials} are \kl{non-negative},
    and such that for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \alpha_0 \prod_{i = 1}^p u_i^{\omega \times X_i} \alpha_i
        \right)
        = P(X_1, \dots, X_p) \quad .
    \end{equation*}
\end{lemma}
\begin{proof}
    Let $\omega$ be an idempotent power for the finite monoid $M$,
    and
    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
    $f = \pi^\dagger$. That is, 
    $f(w)$ is the sum over all factorizations of $w$
    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.

@@

\begin{corollary}
    \label{n-rat-correct:lem}
    Let $P \in \Rel[X_1, \dots, X_p]$ be \kl{represented}
    by a \kl{$\Nat$-rational series}. Then,
    $P \in \CorrectPoly$.
\end{corollary}
\begin{proof}
    Let $f$ be the \kl{commutative}
    \kl{$\Nat$-rational series}
    with domain $\Sigma \defined \set{a_1, \dots, a_p}$
    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
    it is in fact a \kl{$\Nat$-polyregular function}
    (\cref{polynomial-rational-polyreg:fact}).
    Using \cref{n-poly-combinatorics:lem},
    there exists a number $\omega \in \Nat$
    and polynomial $Q$ with \kl{positive coefficients} for its
    \kl{maximal monomials} such that
    for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^p (a_i)^{\omega \times X_i}
        \right)
        = Q(X_1, \dots, X_p)
        = P(\omega X_1, \dots, \omega X_p) 
        \quad .
    \end{equation*}
    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
    as polynomials,
    and in particular that
    the \kl{maximal monomials} of 
    $P$ have \kl{positive coefficients}.

@@

\AP The core of the upcoming \cref{lem:correct-to-n-rat} leverages a notion of
\kl{discrete derivative} to perform an induction on the \kl{maximal monomials},
and suitable notions of \kl{translations} of the domain of polynomials. Namely,
we write $\translate{K}$ for the \intro{translation function} that maps a
polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 + K, \dots,
X_n + K)$.

@@

In the actual statement of \cref{corrected-version:thm}, the representation of
polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
functions} to \kl{star-free $\Nat$-polyregular functions}. This is not
surprising, given the conjecture that \kl{ultimate polynomial}
\kl{$\Nat$-polyregular functions} (re-introduced in \cref{star-free:sec}) were
conjectured to be star free in \cite{LOPEZ23b}.

@@

The following \cref{derivation-simplifies:lemma,derivation-covers:lemma} 
show how the \kl{discrete differentiation} operator
interacts with \kl{maximal monomials} with respect to the \kl{Hoare ordering}:
it extracts \emph{submaximal} monomials from a given polynomial.

@@

\begin{fact}
    Let $K > 0$,
    and let $M,T$ be two \kl{monomials}, such that
    $T$ \kl{strictly divides} $M$.
    Then,
    $T$ \kl{divides} some \kl{maximal monomial}
    of $\Diff{K}{M}$.
\end{fact}

\begin{lemma}
    \label{derivation-covers:lemma}
    For all $P,Q \in \Rel[\vec{X}]$ that are non-constant,
    for all $K > 0$,
    if $\MaximalMonomials(P) \hoarele \MaximalMonomials(Q)$,
    then
    $\MaximalMonomials(P) \hoareleq \MaximalMonomials(\Diff{K}{Q})$.
\end{lemma}

@@

While \cref{lem:correct-to-n-rat} provides effective conversion, it does not
explicitly state that the membership is decidable to keep the proof clearer. A
similar proof scheme can be followed to conclude that membership is decidable.

\begin{lemma}
    \label{decidability-correct:lem}
    It is decidable whether a polynomial 
    $P \in \Rel[\vec{X}]$
    belongs to $\CorrectPoly$.
\end{lemma}
\begin{proof}
    Let $P \in \Rel[\vec{X}]$. We prove by induction on
    $\MaximalMonomials(P)$ that whether $P \in \CorrectPoly$
    is decidable.

### 2024-01-18 parts/counter.tex
re-organising transducers and such

%! TeX program = xelatex
%! lang = en-US
\section{The Counter Example}
\label{sec:c-example}

We will start by formally stating the result of \citeauthor{KARH77} in terms of
polynomials and commutative $\Nat$-rational series, and then introduce the
necessary definitions. The characterization of commutative $\Nat$-rational
series computing polynomial functions is described in terms of a class
$\CoveredPoly[\vec{X}]$ of polynomials with multiple indeterminate.

@@

Let us now introduce the counter example that was found trying to generalize
the original proof to non-commutative \kl{$\Rel$-rational series} of
\kl{polynomial growth}. The counter example will use three indeterminate, and
we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
polynomial with at most two indeterminate.

@@

As we will see in the next section, the characterization of \kl{$\Nat$-rational
polynomials} using the set $\CoveredPoly$ holds when the polynomials have at
most $2$ indeterminate. This shows that the examples in \cite{KARH77} are not
invalidated, as they all use at most two indeterminates. Furthermore, the
converse implication, that is, that \kl{$\Nat$-rational polynomials} belong to
$\CoveredPoly$ is correct, and can even be strengthened as we will see in
\cref{corrected-version:thm}.

>>>>

%! TeX program = xelatex
%! lang = en-US
\section{Polynomials}

The goal of this section is to provide semantic characterizations of
\kl{$\Nat$-rational polynomials}, decide if a polynomial $P$ is a
\kl{$\Nat$-rational polynomial}, and if so, explicitly compute a
\kl{$\Nat$-polyregular function} $f$ that \kl{represents} $P$.

It is an easy check that polynomials with coefficients in $\Nat$ are
\kl{$\Nat$-rational polynomials} (\cref{n-poly-n-poly:example}). However,
\cref{negative-but-npoly:ex} provides an example of a polynomial with negative
coefficients that is a \kl{$\Nat$-rational polynomial}. As a consequence, the
class of interest strictly contains $\Nat[\vec{X}]$.

\begin{lemma}
    \label{n-poly-n-poly:example}
    Let $P \in \Nat[\vec{X}]$. Then, $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{lemma}
\begin{proof}
    We prove it by remarking that $\NPoly$
    is stable under non-negative linear combinations
    and contains the functions ...
\end{proof}

\begin{example}
    \label{negative-but-npoly:ex}
    The polynomial $P(X) \defined X^2 -2X + 2$ is
    a \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    Remark that $P = (X-1)^2 + 1$. Let $f  \colon \Sigma^* \to \Nat$
    that maps $aw \mapsto |w|^2$, and $\varepsilon \to 0$.
    It is clear that $f \in \NPoly$, and that 
    $P$ is \kl{represented} by
    $\ind{|w| \geq 1} \times f + \ind{|w| = 0}$.
\end{proof}

Let us now introduce the characterization of \kl{$\Nat$-rational polynomials}
given by \citeauthor{KARH77}, which is stated using a class
$\CoveredPoly[\vec{X}]$ of polynomials with multiple indeterminate.

@@

\subsection{The counter example}
\label{sec:c-example}

Let us now introduce the counter example that was found trying to generalize
the original proof to non-commutative \kl{$\Rel$-rational series} of
\kl{polynomial growth}. The counter example will use three indeterminate, and
we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
polynomial with at most two indeterminate. In particular, the examples
appearing in \cite{KARH77} are not invalidated, as they all use at most two
indeterminates. Furthermore, the converse implication, that is, that
\kl{$\Nat$-rational polynomials} belong to $\CoveredPoly$ is correct, and can
even be strengthened as we will see in \cref{corrected-version:thm}.

@@

### 2024-01-18 parts/intro.tex
re-organising transducers and such

There have been recurrent attempts to generalize the notions of regularity from
languages to functions, leading to classes of various expressiveness
(sequential, rational, regular, and \kl{polyregular functions}) \cite[see,
e.g.][]{bojanczyk2019string}. In this paper, we will focus on \kl{polyregular
functions}, that shares with regular languages the characterizations in terms
of logic (via $\MSO$ interpretations), automata (via pebble automata or SST
\cite{}), and adds new bridges with programming languages (lambda-terms)
for-transducers. Although the class of "star-free" polyregular functions is
defined and similarly characterized using first-order logic.

Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton.
Indeed, for sequential functions that admit a canonical object, aperiodicity is
decidable \cite{choffrut03}, and more recently, it aperiodicity has been proven
decidable for rational functions
\cite{filiot2016aperiodicity,filiot2018canonical}. The problem remains open for
regular functions, although it has been proven decidable when changing the
semantics of function equality to account for the \emph{origins} in the output
\cite{bojanczyk14}.

Another way to approach the problem of deciding star-free for polyregular
functions is to not restrict the computational power, nor change the semantics,
but consider restrictions of the output alphabet. In particular, there have
been successful results proven on \emph{unary polyregular functions}, also
known as \kl{$\Nat$-polyregular functions}
\cite{doueneau2021pebble,doueneau2022hiding}. Let us note that results on
\emph{unary output functions} often generalize to the slightly more general
class of functions with \emph{commutative output}, and that algorithms leverage
the absence of combinatorial information in the output. Furthermore,
\kl{$\Nat$-polyregular functions} benefit from theorems proven in the field of
rational series \cite{berstel2011noncommutative}, where they coincide with
weighted automata of polynomial ambiguity, rational series of polynomial
growth, and a notion of quantitative $\MSO$
\cite{berstel2011noncommutative,kreutzer2013,schutzenberger1962}. Although
promising, the restriction to unary output did not yield effective procedure to
decide the aperiodicity of polyregular functions.
Notice that unary outputs in $\Rel$ and $\Nat$
are therefore special kind of weighted automata,
and that questions arising from the weighted automata
theory (good first order logic?) are also connected.
\cite{droste2019aperiodic,reutenauer_series_1980}.

Recently, the introduction of $\Rel$-output polyregular functions, called
\kl{$\Rel$-polyregular functions} (and more generally, commutative invertible
outputs) allowed to solve the general problem \cite{LOPEZ23b}. Note that
$\Rel$-output generalizes $\Nat$-output, but somehow has a combinatorial
description that is farther away from traditional \kl{polyregular functions}
because of the invertible outputs. The proof relies on the construction of a
canonical object essentially based on differences between functions, which is
not applicable to unary output. Note that already for rational series, the
problem is open.

We write $\Poly$ for the class of \kl{polyregular functions},
$\SF$ for the class of \kl{star-free polyregular functions},
$\NRat$ for the class of \kl{$\Nat$-rational series},
and $\ZRat$ for the class of \kl{$\Rel$-rational series}.
Arrows denote strict inclusions, and effectiveness
(both in terms of decidability and of effective representation)
is represented by thick double arrows.
Inclusions that are suspected to be effective have a question mark.
It is also conjectured that $\NPoly \cap \ZSF = \NSF$.

\begin{figure}
    \includestandalone[width=\linewidth]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions. 
    }
\end{figure}

It was conjectured in a thesis that some particular form of polynomials
would characterize star-free functions.This paper follows the line of
research on \kl{polyregular functions} that focuses on semantic
characterizations of the asymptotic behavior, rather than the construction
of canonical objects of computation
\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}. The goal was to
decide arrows X,Y and Z, namely, to understand precisely the relationship
between star-free and $\Nat$ output. In the case of \kl{$\Nat$-rational
series}, a result of \citeauthor{KARH77} characterizes \kl{commutative}
\kl{$\Nat$-rational series} among \kl{$\Rel$-rational series}
\cite{KARH77}.
This conjecture was based on the previously introduced result of
\textcite[Theorem 3.3]{KARH77}, which we will disprove in \cref{sec:c-example}.
We then provide a corrected version of the theorem, and prove the conjecture of
Douéneau in the case of \emph{commutative} $\Nat$-rational series of polynomial
growth.

@@

\paragraph{Contributions.} 
\begin{itemize}
    \item We provide a candidate semantic characterization
    \item We prove that it is correct for commutative inputs
    \item We provide a counter-example to the theorem of nanania.
    \item We show that effective procedures are possible for all arrows 
        and answer positively to the conjecture of XX
        in the case of commutative inputs.
\end{itemize}

\paragraph*{Outline of the paper.}
\begin{itemize}
    \item In a first section, we recall the preliminary definitions.
    \item In a second section, we provide a counter example to the
        theorem of XX.
    \item In a third section, we provide an alternative result.
    \item In a fourth section, we make the results effective,
        and compute everyting.
\end{itemize}

>>>>

There have been recurrent attempts to generalize the notions of regularity from
languages to functions, leading to classes of various expressiveness
(sequential, rational, regular, and \kl{polyregular functions}) \cite[see,
e.g.][]{bojanczyk2019string}. In this paper, we will focus on \kl{polyregular
functions}, whose class is denoted $\Poly$, and that shares with regular
languages the characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers. As for
regular languages, the class $\SF$ of \intro{star-free polyregular function}
can equivalently be introduced in terms of first-order definability,
counter-free pebble transducers, and suitable restrictions of for-tranducers.

Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton.
Indeed, for sequential functions that admit a canonical object, aperiodicity is
decidable \cite{choffrut03}, and more recently, aperiodicity has been proven
decidable for rational functions
\cite{filiot2016aperiodicity,filiot2018canonical}. The problem remains open for
regular functions, although it has been proven decidable when changing the
semantics of function equality to account for the \emph{origins} in the output
\cite{bojanczyk14}.

Another way to approach the problem of deciding star-free for polyregular
functions is to not restrict the computational power, nor change the semantics,
but consider restrictions of the output alphabet. In particular, there have
been successful results proven on \emph{unary polyregular functions}, also
known as \kl{$\Nat$-polyregular functions}
\cite{doueneau2021pebble,doueneau2022hiding}. Let us note that results on
\emph{unary output functions} often generalize to the slightly more general
class of functions with \emph{commutative output}, and that algorithms leverage
the absence of combinatorial information in the output. Furthermore,
\kl{$\Nat$-polyregular functions} benefit from theorems proven in the field of
rational series \cite{berstel2011noncommutative}, where they coincide with
weighted automata of polynomial ambiguity, rational series of polynomial
growth, and a notion of quantitative $\MSO$
\cite{berstel2011noncommutative,kreutzer2013,schutzenberger1962}. Although
promising, the restriction to unary output did not yield effective procedure to
decide the aperiodicity of polyregular functions.

Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} called \kl{$\Rel$-polyregular functions} and denoted
by $\ZPoly$ \cite{LOPEZ23b}.  These functions generalise \kl{$\Nat$-polyregular
functions} by allowing negative outputs, and therefore enabling a limited form
of backtracking. The decidability of aperiodicity for \kl{$\Rel$-polyregular
functions} relies on the construction of a canonical object called the
\intro{residual transducer}, the latter being essentially based on differences
between functions, crucially leveraging negative outputs.
If the proof method does not carry from $\Rel$-output to $\Nat$-output
functions, it was conjectured that the semantic characterization of \kl{star
free $\Rel$-polyregular functions} ($\ZSF$) would continue to characterize
\kl{star-free $\Nat$-polyregular functions} ($\NSF$), and in particular that
$\NPoly \cap \ZSF = \NSF$.

Let us remark that, \kl{$\Nat$-polyregular functions} and
\kl{$\Rel$-polyregular functions} are proper subsets respectively of
\kl{$\Nat$-rational series} ($\ZRat$) and of \kl{$\Rel$-rational series}
($\NRat$). For these larger classes of functions, several competing notions of
aperiodicity exist \cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b},
and decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ among $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved in the case
of \emph{polynomials} by \textcite{KARH77}.

We refer the reader to \cref{previously-known-inclusions:fig} to get an
overview of the different classes of functions appearing in the paper. Arrows
denote strict inclusions, and effectiveness (both in terms of decidability and
of effective representation) is represented by thick double arrows. Inclusions
that are suspected to be effective are represented using a dashed arrow
together with a question mark.

\begin{figure}
    \includestandalone[width=\linewidth]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Nat$, $\Rat$, $\Sigma$). The second
        one is the allowed computational power
        (star-free polyregular functions, polyregular functions, 
        rational series).
    }
    \label{previously-known-inclusions:fig}
\end{figure}

\paragraph{Contributions.} We prove that for \kl{commutative} functions, all
the membership problems involving the classes $\ZSF$, $\ZPoly$, $\NSF$, and
$\NPoly$ are all decidable (this is the lower left square of inclusions in
\cref{previously-known-inclusions:fig}), and effective conversion algorithms
are provided. Furthermore, we show that in the commutative case, the conjecture
stating that $\ZSF \cap \NPoly = \NSF$ holds.

These results are obtained by completely understanding which \emph{polynomials}
are computable using \kl{$\Nat$-polyregular functions}, and lifting these
results from polynomials to \kl{$\Rel$-polyregular functions}. This study
invalidates the main theorem of \cite{KARH77}, for which we provide an explicit
counter-example.

\begin{itemize}
    \item We show that commutativity is decidable.
    \item We provide a candidate semantic characterization
    \item We prove that it is correct for commutative inputs
    \item We provide a counter-example to the theorem of nanania.
    \item We show that effective procedures are possible for all arrows 
        and answer positively to the conjecture of XX
        in the case of commutative inputs.
\end{itemize}
It was conjectured in a thesis that some particular form of polynomials
would characterize star-free functions.

This paper follows the line of
research on \kl{polyregular functions} that focuses on semantic
characterizations of the asymptotic behavior, rather than the construction
of canonical objects of computation
\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}. The goal was to
decide arrows X,Y and Z, namely, to understand precisely the relationship
between star-free and $\Nat$ output. In the case of \kl{$\Nat$-rational
series}, a result of \citeauthor{KARH77} characterizes \kl{commutative}
\kl{$\Nat$-rational series} among \kl{$\Rel$-rational series}
\cite{KARH77}.
This conjecture was based on the previously introduced result of
\textcite[Theorem 3.3]{KARH77}, which we will disprove in \cref{sec:c-example}.
We then provide a corrected version of the theorem, and prove the conjecture of
Douéneau in the case of \emph{commutative} $\Nat$-rational series of polynomial
growth.

\cref{thm:counter-example}
\cref{corrected-version:thm}

\cref{decidable-n-poly:thm}
\cref{zsf-npoly-nsf:thm}

@@

\paragraph*{Outline of the paper.}
In \cref{preliminaries:sec}, we introduce the computational models
and prove that 
\begin{itemize}
    \item In a first section, we recall the preliminary definitions.
    \item In a second section, we provide a counter example to the
        theorem of XX.
    \item In a third section, we provide an alternative result.
    \item In a fourth section, we make the results effective,
        and compute everyting.
\end{itemize}

### 2024-01-18 parts/prelim.tex
re-organising transducers and such

>>>>

Let us briefly remark that the characterizations of subclasses of
\kl{polyregular functions} that are obtained in practice can be obtained by
pre-composition with a \kl{commutative} \kl{regular function}.

\begin{remark}
    Let $f \in \ZRat$, and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item $f \in \ZPoly[k]$,
        \item For every \kl{commutative} \kl{polyregular function} $h$
            of \kl{growth rate} $l \in \Nat$,
            $(f \circ h) \in \ZPoly[k+l-1]$.
    \end{enumerate}
\end{remark}

\begin{remark}
    Let $f \in \ZPoly$.
    The following are equivalent:
    \begin{enumerate}
        \item $f \in \ZSF$,
        \item For every \kl{commutative} \kl{star-free polyregular function} $h$,
            $(f \circ h) \in \ZSF$.
    \end{enumerate}
\end{remark}

### 2024-01-18 parts/starfree.tex
re-organising transducers and such

@@

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and 
    let $f \in \ZPoly[k]$ be a \kl{commutative} \kl{$\Rel$-polyregular function}.
    The following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item $f \in \NPoly$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and one can effectively compute a representation of $f$.
\end{theorem}

To prove the theorem, let us first provide a nice characterization of
\kl{commutative} \kl{$\Rel$-polyregular functions} in terms of polynomials.
This is actually a refinement of \cref{n-poly-combinatorics:lem} in the
commutative case. Informally, we state that \kl{commutative}
\kl{$\Rel$-polyregular functions} are obtained by applying different
\kl{$\Rel$-rational polynomials} based on the modulos...

@@

>>>>

Let us now move on from polynomials to general \kl{$\Rel$-polyregular
functions}. As we will prove, \kl{commutativity} reduces $\ZPoly$ to a
combination of a finite number of polynomials, arranged using modulo counting.
We will first decide $\NPoly$ inside $\ZPoly$, and then leverage this to decide
$\NSF$ inside $\NPoly$, incidentally proving that $\ZSF \cap \NPoly = \NSF$.

@@

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and 
    let $f \in \ZPoly[k]$ be a \kl{commutative} \kl{$\Rel$-polyregular function}.
    The following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and one can effectively compute a representation of $f$.
\end{theorem}

To prove the theorem, let us first provide a nice characterization of
\kl{commutative} \kl{$\Rel$-polyregular functions} in terms of polynomials.
This is actually a refinement of \cref{n-poly-combinatorics:lem} in the
commutative case.

@@

\begin{definition}
    Partial applications $\app{f}{u} \defined w \mapsto f(uw)$.
\end{definition}

\begin{definition}
    Residuals $\Res(f) \defined \setof{ \app{f}{u} }{u \in \Sigma^*}$.
\end{definition}
\begin{definition}
    We say that $f \natleq[k] g$ if and only if $g - f \in \NPoly[k]$.
\end{definition}
